---
title: "Probability Practice"
author: " "
date: "`r Sys.Date()`"
output: pdf_document
---

## Probability Practice

```{r packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidyr)
library(corrplot)
library(dbscan)
```

**Part A) Visitors to your website are asked to answer a single survey question before they get access to the content on the page. Among all of the users, there are two categories: Random Clicker (RC), and Truthful Clicker (TC). There are two possible answers to the survey: yes and no. Random clickers would click either one with equal probability. You are also giving the information that the expected fraction of random clickers is 0.3. After a trial period, you get the following survey results: 65% said Yes and 35% said No. What fraction of people who are truthful clickers answered yes? Hint: use the rule of total probability.** \newline

Y = yes \newline
N = no \newline
RC = random clicker (1-B) \newline
TC = truthful clicker (B) \newline

P(Y | RC) = 0.50 \newline
P(N | RC) = 0.50 \newline 

Fraction of random clickers is 0.30 so: \newline
P(RC) = 0.30\newline
P(TC) = 1 - P(RC) = 1.0 - 0.30 = 0.70 \newline

After trial period: \newline
P(Y) = 0.65
P(N) = 0.35

By total probability... \newline
P(A) = P(A|B)P(B) + P(A|1-B)P(1-B) -> P(Y) = P(Y|TC)P(TC) + P(Y|RC)P(RC) \newline
P(A|B) = P(A) - ( P(A|1-B)P(1-B) ) / P(B)

```{r probabilityA}
p_RC <- 0.3 
p_TC <- 1 - p_RC

p_Yes_RC <- 0.5
p_Yes_total <- 0.65 

p_Yes_TC <- (p_Yes_total - (p_Yes_RC * p_RC)) / p_TC
p_Yes_TC
```

**Answer:** Using the total probability rule we find that the fraction of people who are truthful clickers that answered yes is 0.71428 or about 71.43% \newline


**Part B) Imagine a medical test for a disease with the following two attributes:** \newline
- The sensitivity is about 0.993. That is, if someone has the disease, there is a probability of 0.993 that they will test positive. \newline
- The specificity is about 0.9999. This means that if someone doesn't have the disease, there is probability of 0.9999 that they will test negative. \newline
- In the general population, incidence of the disease is reasonably rare: about 0.0025% of all people have it (or 0.000025 as a decimal probability). \newline

**Suppose someone tests positive. What is the probability that they have the disease?** \newline

D = Someone that has the disease \newline
H = Someone that does **not** have the disease \newline
P = Testing positive \newline
N = Testing negative \newline

P(D) = 0.000025 \newline
P(H) = 1 - 0.000025 = 0.99975 \newline
P(P|D) = 0.993 \newline
P(N|H) = 0.9999 \newline

Supposing someone tests positive, what is the probability that they have the disease can be represented by P(D|P) \newline

First: \newline

P(N|D) = 1 - 0.993 = 0.007 \newline
P(P|H) = 1 - 0.9999 = 0.0001 \newline

The probability that someone has the disease given that they test positive *( D|P )* is represented by: \newline

**P(D|P) = ( P(P|D)P(D) ) / P(P)** \newline

To calculated this we need to find the probability that someone tests positive. The probability that a person tests positive can be calculated by: \newline

P(P) = P(P|D)P(D) + P(P|H)P(H) \newline
P(P|D) = 0.993 \newline
P(P|H) = 1 - 0.9999 = 0.0001 \newline
P(D) = 0.000025 \newline
P(H) = 1 - 0.000025 = 0.999975 \newline

**P(P)** = (0.993)(0.000025) + (0.0001)(0.999975) = 0.0001248225 \newline

Now to calculated P(D|P): \newline

P(D|P) = (0.993*0.000025)/0.0001248225


```{r probabilityB}

sensitivity <- 0.993  #p(p|d)
specificity <- 0.9999 #p(n|h)
prevalance <- 0.000025 #p(d)


false_positive <- 1 - specificity #p(p|h)

prob_positive <- (sensitivity*prevalance) +(false_positive*(1-prevalance))

p_disease_g_positive <- (sensitivity*prevalance)/prob_positive
p_disease_g_positive
```

**Answer:** The probability of having the disease given that they tested positive is 0.19882 or 19.89%